---
title: "Intro to R and Data Visualization with ggplot2: Session 1 part 2: intro EDA with R"
subtitle: UCI Data Science Initiative
date: "22 June, 2020"
author: "Arnold Seong"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css:
      - default
      - default-fonts
      - "https://cdnjs.cloudflare.com/ajax/libs/animate.css/3.7.0/animate.min.css"
      - xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---



```{r setup, include=FALSE, message=F, echo=F, warning=F}
library(xaringanthemer)
style_mono_accent(base_color = "#3C989E")
knitr::opts_chunk$set(cache=FALSE, 
                      warning=FALSE, 
                      message=FALSE, 
                      echo=TRUE, 
                      warning=FALSE, 
                      error=FALSE)
options(scipen=10)
library(ggplot2)
library(RColorBrewer)
library(kableExtra)
```

layout: true

<style>
pre {
    display: block;
    font-family: monospace;
    white-space: pre;
    margin: 1em 0px;
    margin-top: 0.5em;
    margin-right: 0px;
    margin-bottom: 0em;
    margin-left: 0px;
}
.remark-inline-code {
  background: #F5F5F5; 
}
</style>




---
class: animated, fadeIn

## Workshop info
+ We will be recording today's sessions for potential future use
    + You may not want to activate video
    + Please set your zoom to mute your audio until you would like to speak
    + You can change your icon name by mousing over, clicking on the 3 dots
    
+ Please ask questions during lectures & throughout the day!
    + Use chat
    + Raise hand (`Alt+Y`)
    + Unmute yourself temporarily (hold `Spacebar`).







---
## Importance of EDA: Anscombe's quartet

**Anscombe's quartet** is a famous set of four (x,y) pairings that illustrate the importance of plotting your data and EDA in general.

- The most common numerical summaries (mean, standard deviation, correlation) are identical

- The linear regression coefficients (regressing y on x) are also identical!

--
#### &nbsp;

```{r, echo=F, warning=F, message=F}
anscombe2 <- with(anscombe, data.frame(
  x     = c(x1, x2, x3, x4),
  y     = c(y1, y2, y3, y4),
  set = gl(4, nrow(anscombe))
))


library(plyr)
stats <- ddply(anscombe2, .(set), summarize, 
  mean_x = mean(x),
  mean_y = mean(y), 
  sd_x = sd(x),
  sd_y = sd(y),
  corr_xy = cor(x, y), 
  intercept = lm(y ~ x)$coefficients[1], 
  x_effect = lm(y ~ x)$coefficients[2]
)
stats[,1] <- as.numeric(stats[,1])

colnames(stats) <- c("set", 
                     "mean(x)", 
                     "mean(y)", 
                     "sd(x)", 
                     "sd(y)", 
                     "cor(x,y)", 
                     "intercept", 
                     "x_effect")


library(kableExtra)

kable(round(stats, 4), caption = "Anscombe's Quartet, rounded to 4 decimals") %>%
  kable_styling(bootstrap_options = c( "striped", full_width = F))
```





---
## Anscombe's Quartet, plotted

```{r, echo=F, warning=F, message=F}
ggplot(anscombe2, aes(x, y)) +
  geom_point(col="hotpink", size=2) +
  facet_wrap(~ set)
```




---
class: inverse
## Session 2 Topics

1. Basic data exploration
  + Getting a sense of your data: summaries, tables
  + Making your data useful: transformations
  + plotting with base R



# EDIT EDIT EDIT



---
## Intro: basic EDA

There is no set process for Exploratory Data Analysis, though there are some guiding principles.

In this portion of the workshop, we'll cover a few things that are almost always good starting points

  - We'll use a commonly-used dataset that comes loaded with R, the `airquality` dataset.
  
  - Our EDA process will become more complex and specific to our NYTimes dataset in the next portions of the workshop.




---
## EDA goals:

In general, there are 2 main goals of EDA:

1. Assess the structure and quality of your data
  
1. Investigate patterns

--

  - variation in 1 variable

--
  
  - covariation between variables
  






---
## Basic EDA: `str()`

A good place to start is knowing something about your dataset!

  - First, look at the help file for `airquality` by typing `?airquality`

After that, try out `str()`, which explains the internal structure of an object.

--

```{r}
str(airquality)
```

--

- What do you notice?

--

- you'll use this more in the exercises!





---
## Basic EDA: `Month` into factor

For illustrative purposes, since we don't have any categorical variables here, let's make a factor.  We'll split the months into "Summer" and "Fall"

- "Summer" will correspond to `Month` values 5, 6, 7

- "Fall" will correspond to `Month` values 8, 9

--

```{r}
airquality$season <- factor(airquality$Month, levels=5:9,
    labels=c("Summer", "Summer", "Summer", "Fall", "Fall"))
```





---
## Basic EDA: `summary()`
A quick first step in any data analysis is to generate what are commonly known as 5-number summaries.  
- You can do this (well, a 6-number version) quickly in R using `summary()`

```{r}
summary(airquality)
```




---
## Basic EDA: `summary()`
```{r, echo=F}
summary(airquality)
```

- For quantitative variables, this gives you a rough sense of the range of your variables, skewness, and presence of extreme observations / data entry errors.

- For qualitative variables, you can see how balanced your dataset is





---
## Did you notice the missingness?

Assessing the amount of missingness in data is crucial, and should be one of your first steps.  

--

- If you didn't see it in the `summary()` slide, here's a different way to see it quickly:

```{r}
apply(airquality, 2, function(x) sum(is.na(x)))
```

--

We see that the missingness is concentrated in `Ozone` and `Solar.R`.  This might be problematic.

--

- The best way to tell if missingness will be problematic is to **hire a statistician** (non-ignorable missingness will bias your results!)

--

- We also know that there are `r nrow(airquality)` observations total.  That's a high rate of missingness in `Ozone` and likely to pose problems.







---
## Basic EDA: Table 1

In many disciplines, the first table in a paper is a description of population characteristics, broken down (stratified) by a categorical variable of interest.  

--

- Let's pretend we're interested in `season` 

--

- we could make several subsets of our data and use `summary()` and format it and...

--


... Or we could use a package.  Turns out there are table-1-generating functions in many packages.  The package `table1` seems pretty good.

```{r, eval=F}
install.packages("table1")
```




---
## Basic EDA: Table 1


```{r, warning=F, message=F}
library(table1)
table1(~Ozone + Solar.R + Wind + Temp + Day | season, data=airquality)
```






---
## Pair plot

For a fast birds-eye view, use `pairs()`.  You can call `pairs()` on your entire dataframe if you wish, but it is not usually very informative for categorical variables.

```{r, fig.width=7, fig.height=5}
pairs(airquality[, c("Ozone", "Solar.R", "Wind", "Temp")])
```




---
## Pair plot

Let's change some plot options so we can see things a little better.

```{r, fig.width=7, fig.height=5}
palette(c("hotpink", "deepskyblue"))
pairs(airquality[, c("Ozone", "Solar.R", "Wind", "Temp")], 
      pch=20, col=airquality$season,
      main="Summer (pink), Fall (blue)")
```






---
## Plotting: `hist()`

Histograms are used to visualize the distribution of quantitative variables.

--

Let's look at the distribution of `Wind`

```{r, fig.width=7, fig.height=5}
hist(airquality$Wind)
lines(density(airquality$Wind))
```




---
## Plotting: `hist()`

Be careful of how your bin width affects your plot.

Here's the same histogram, with smaller bin width (changed using `breaks=` argument).

```{r, fig.width=7, fig.height=5}
hist(airquality$Wind, breaks=20)
```




---
## Plotting: `hist()` and `density()`

Setting `freq=FALSE` in `hist()` and adding a line for the density can help.

```{r, eval=F}
hist(airquality$Wind, breaks=20, freq=FALSE)
lines(density(airquality$Wind))
```


- `freq=FALSE` changes the y-axis to percentages of total observations rather than counts

- `density()` performs a kernel density estimation (essentially a smoothed histogram)

- `lines()` adds lines to the current plot





---
## Plotting: `hist()` and `density()`

Setting `freq=FALSE` in `hist()` and adding a line for the density can help.

```{r, fig.width=7, fig.height=5}
hist(airquality$Wind, breaks=20, freq=FALSE)
lines(density(airquality$Wind))
```





---
## Plotting: modifying the plot

```{r, fig.width=7, fig.height=5}
hist(airquality$Wind, breaks=20, freq=FALSE,
     main="", ylab="Density", xlab="Wind (mph)",
     col="lightgrey")
lines(density(airquality$Wind), 
      col="hotpink", lwd=2)
```






---
## Plotting: boxplots

Like histograms, boxplots also display information about a variable's density.

- Less detail, but more emphasis on numeric summary statistics (quartiles)

Use `boxplot()` in R

```{r, fig.width=7, fig.height=5}
boxplot(airquality$Solar.R, horizontal=TRUE, xlab="Solar Rad. (lang)")
```





---
## Plotting: grouped boxplots

Used for 1 quantitative and 1 qualitative variable

```{r, fig.width=7, fig.height=5}
boxplot(airquality$Solar.R~airquality$season, ylab="Solar Rad. (lang)", xlab="Season")
```



---
## Plotting: grouped boxplots

Let's look at this against `Month`

```{r, fig.width=7, fig.height=5}
airquality$month_fac <- factor(airquality$Month, levels=5:9,
                  labels=c("May", "Jun", "Jul", "Aug", "Sep"))
boxplot(airquality$Solar.R~airquality$month_fac, ylab="Solar Rad. (lang)", xlab="Month")
```






---
## Plotting: Scatterplots

- Used for 2 quantitative variables
- Use `plot()`

```{r, fig.width=7, fig.height=5}
plot(airquality$Ozone~airquality$Wind)
```





---
## Plotting: Scatterplots

Let's make it prettier, use color to indicate factor levels, and add a legend.

```{r, fig.width=6, fig.height=4}
pal2 <- c("hotpink", "deepskyblue")
palette(pal2)
plot(airquality$Ozone~airquality$Wind, col=airquality$season, pch=19, 
     main="Ozone vs Wind",xlab="Wind (mph)", ylab="Ozone(ppb)")
legend("topright", bty="n", pch=19, col=pal2,
       legend=levels(airquality$season))

```







---
## NEXT:

- Exercises 1 during break; review afterwards
- Next sessions will be about ggplot2

