---
title: "Intro to R and Data Visualization with ggplot2: Session 2 part 1"
subtitle: UCI Data Science Initiative
params: 
  reportdate: "`r format(Sys.time(), '%d %B %Y')`"  # to be replaced
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "Arnold Seong"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css:
      - default
      - default-fonts
      - "https://cdnjs.cloudflare.com/ajax/libs/animate.css/3.7.0/animate.min.css"
      - xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

  
# output:
#   html_document:
#     highlight: tango

---



```{r setup, include=FALSE, message=F, echo=F, warning=F}
library(xaringanthemer)
# style_xaringan(
  # code_inline_background_color = "#C0C0C0"
#   text_font_family = "Droid Serif",
#   text_font_url = "https://fonts.googleapis.com/css?family=Droid+Serif:400,700,400italic",
#   header_font_google = google_font("Yanone Kaffeesatz")
# )
style_mono_accent(base_color = "#3C989E")

# knitr::opts_chunk$set(cache=FALSE, warning=FALSE, message=FALSE, echo=TRUE, warning=FALSE, error=FALSE)
knitr::opts_chunk$set(cache=TRUE, warning=TRUE, message=TRUE, echo=TRUE, warning=TRUE, error=TRUE)

#set scientific notation digit threshold
options(scipen=10)

#load libraries
library(knitr)
library(ggplot2)
library(RColorBrewer)
```

layout: true

<style>
pre {
    display: block;
    font-family: monospace;
    white-space: pre;
    margin: 1em 0px;
    margin-top: 0.5em;
    margin-right: 0px;
    margin-bottom: 0em;
    margin-left: 0px;
}
.remark-inline-code {
  background: #F5F5F5; 
}
</style>


<!-- xaringan::inf_mr() -->

---
class: animated, fadeIn

## Workshop info
+ We will be recording today's sessions for potential future use
    + You may not want to activate video
    + Please set your zoom to mute your audio until you would like to speak
    + You can change your icon name by mousing over, clicking on the 3 dots
    
+ Please ask questions during lectures & throughout the day!
    + Use chat
    + Raise hand (`Alt+Y`)
    + Unmute yourself temporarily (hold `Spacebar`).


<!-- + To access course materials please visit the [IDA-with-R website](https://ucidatascienceinitiative.github.io/IDA-with-R/). -->
<!-- + **Please [download](https://github.com/UCIDataScienceInitiative/IDA-with-R/archive/master.zip) & unzip the Github repository!** -->


---
class: inverse
## Session 2 Topics

1.  Introduction to ggplot2 and the "grammar of graphics"
  - understanding the logic of ggplot2
  - preparing and plotting data.
1. Beyond ggplot2 basics
  - overlaying geoms
  - visualizing strata for comparison
  - changing default plot options.




---
class: inverse, animated, fadeIn
## Session 2 Topics

1.  Introduction to ggplot2 and the "grammar of graphics"
  - understanding the logic of ggplot2
  - preparing and plotting data.
1. **Beyond ggplot2 basics**
  - **overlaying geoms**
  - **visualizing strata for comparison**
  - **changing default plot options.**






---
## Intro to ggplot2

- `ggplot2` is an R package for statistics and data visualization 

- developed by Hadley Wickham, based on the Grammar of Graphics (Wilkinson 2005)

--

- underlying principle: just like a sentence, a graph is made up of independent components that can be combined according to general rules.


--

- further resources:
  - [2010 paper by Hadley Wickham](https://vita.had.co.nz/papers/layered-grammar.html) describing ggplot2
  - [ggplot2 online book](https://ggplot2-book.org/), also by Wickham

- Much of this portion will draw on these two resources!




---
## `ggplot2` advantages / disadvantages

Advantages:
  - modularity and iterative updating
  - cohesive framework
  - creativity - move beyond named graphics


Disadvantages:
  - just as easily nonsensical
  - like learning a new language sometimes
      - in terms of the "grammar of graphics"
      - neologisms - i.e. what is a "grob"?
  - data must be provided in a certain form




---
## `ggplot2()` advantages in practice

Advantages:

  - Particularly helpful for complex plotting tasks:
  
    - Many issues regarding graph clarity (e.g. legends) are taken care of
  
    - Modularity allows us to more easily build, iterate, rebuild plots layer by layer, mapping by mapping
  
--

  - Prettier out of the box
  
--

  - More development taking place.  E.g. [`plotly`](https://plotly.com/ggplot2/)



--

You can do in base R anything you can do in ggplot2, and vice versa.  Some things are easier in base R, some in ggplot2 ... and a lot of this depends on your familiarity with each one.




---
## Logic of ggplot2 
(adapted from [ggplot2-book](https://ggplot2-book.org/getting-started.html#introduction-1))

Three key pieces to defining a ggplot2 plot:

1. **data**

  - must be a dataframe
  - use long format

--

1. a set of **aesthetic mappings** between variables in the data and visual properties

--

  - for example: 
      - frequency --> y-position
      - binned values --> x-position
      - factor level --> color
      
--

1. a layer (usually a **geom**) which describes how to render each observation

--

  - continuing the example:
      - layer: bars




---
## ggplot2 syntax

In general, we declare our data and aesthetic mapping using `ggplot()`, then "add" a geometric object layer.

Many different geometric layers exist.  Most common are
- `geom_point()`, `geom_smooth()`
- `geom_boxplot()`
- `geom_histogram()`, `geom_freqpoly()`, `geom_density()`
- `geom_line()`
- `geom_bar()`




---
## ggplot2: example dataset: `diamonds`

- using `diamonds` dataset (included in `ggplot2` package)

```{r, echo=F}
knitr::kable(head(diamonds), format = 'html')
```

- we'll make some scatterplots with this data of `price` vs `carat`, using `color` or `cut` as factors.



---
## ggplot2: scatterplot


```{r, eval=FALSE, fig.width=7, fig.height=5}
ggplot(data=diamonds, aes(x=carat, y=price)) + 
  geom_point()
```


--

```{r, echo=FALSE, fig.width=7, fig.height=5}
ggplot(data=diamonds, aes(x=carat, y=price)) + 
  geom_point()
```





---
## ggplot2: scatterplot


```{r, eval=FALSE, fig.width=7, fig.height=5}
ggplot(data=diamonds, aes(x=carat, y=price, color=color)) + 
  geom_point()
```


--

```{r, echo=FALSE, fig.width=7, fig.height=5}
ggplot(data=diamonds, aes(x=carat, y=price, color=color)) + 
  geom_point()
```





---
## ggplot2: scatterplot + smooth


```{r, eval=FALSE, fig.width=7, fig.height=5}
ggplot(data=diamonds, aes(x=carat, y=price, color=color)) + 
  geom_point(alpha=.25) +
  geom_smooth()
```


--

```{r, echo=FALSE, fig.width=7, fig.height=5}
ggplot(data=diamonds, aes(x=carat, y=price, color=color)) + 
  geom_point(alpha=.25) + 
  geom_smooth()
```



---
## syntax

```{r, eval=F}
ggplot(data=diamonds, aes(x=carat, y=price, color=color)) + 
  geom_point(alpha=.25) + 
  geom_smooth()
```


- First line: global properties
  - all geoms use the data and mapping defined here unless otherwise specified
  
- Second line: add a point layer (i.e. data + mapping displayed as points)
  - added transparency to this layer using `alpha=.25` 

- 3rd line: add a smooth layer (data + mapping processed with a smoother and displayed as a line)
  - default smooth method is a Generalized Additive Model with penalized cubic regression splines.




---
## changed color to cut

base R would require rewriting the legend...

```{r, fig.width=7, fig.height=5, message=FALSE}
ggplot(data=diamonds, aes(x=carat, y=price, color=cut)) + 
  geom_point(alpha=.25) + 
  geom_smooth()
```




---
## `color` aesthetic in point layer only

```{r, eval=F}
ggplot(data=diamonds, aes(x=carat, y=price)) + 
  geom_point(aes(color=cut), alpha=.25) + 
  geom_smooth()
```

- now we will only get one smooth line (using all the data).

--

```{r, echo=F, fig.width=7, fig.height=5, message=F}
ggplot(data=diamonds, aes(x=carat, y=price)) + 
  geom_point(aes(color=cut), alpha=.25) + 
  geom_smooth()
```





---
## Grouped histograms?

```{r, eval=F, fig.width=7, fig.height=5}
ggplot(data=diamonds, aes(x=price)) + 
  geom_histogram(aes(color=cut), alpha=.25)
```

--

Hmmmm, no....  (also, notice the auto-generated message)

```{r, echo=F, fig.width=7, fig.height=5}
ggplot(data=diamonds, aes(x=price)) + 
  geom_histogram(aes(color=cut), alpha=.25)
```






---
## Grouped histogram: attempt 2

```{r, eval=F, fig.width=7, fig.height=5, message=F}
ggplot(data=diamonds, aes(x=price, color=cut, fill=cut)) + 
  geom_histogram(alpha=.25)
```

--

Hmmmm, prettier but still no....

```{r, echo=F, fig.width=7, fig.height=5, message=F}
ggplot(data=diamonds, aes(x=price, color=cut, fill=cut)) + 
  geom_histogram(alpha=.25)
```




---
## Grouped histogram: attempt 3

```{r, eval=F, fig.width=7, fig.height=5, message=F}
ggplot(data=diamonds, aes(x=price, color=cut, fill=cut)) + 
  geom_histogram(alpha=.25, position="identity")
```

--

OK!  but they all overlap, so we can't see anything

```{r, echo=F, fig.width=7, fig.height=5, message=F}
ggplot(data=diamonds, aes(x=price, color=cut, fill=cut)) + 
  geom_histogram(alpha=.25, position="identity")
```
 



---
## Grouped histograms: facet-wrapped

```{r, eval=F, fig.width=7, fig.height=5, message=F}
ggplot(data=diamonds, aes(x=price, color=cut, fill=cut)) + 
  geom_histogram(alpha=.25) + 
  facet_wrap(~cut)
```

--

better! But we have too many "Ideal" diamonds (y-axis scaling poor)

```{r, echo=F, fig.width=7, fig.height=5, message=F}
ggplot(data=diamonds, aes(x=price, color=cut, fill=cut)) + 
  geom_histogram(alpha=.25) + 
  facet_wrap(~cut)
```
 



---
## Grouped histograms: facet-wrapped

```{r, eval=F, fig.width=7, fig.height=5, message=F}
ggplot(data=diamonds, aes(x=price, y=..density.., 
                          color=cut, fill=cut)) + 
  geom_histogram(alpha=.25) + 
  facet_wrap(~cut)
```

--

```{r, echo=F, fig.width=7, fig.height=5, message=F}
ggplot(data=diamonds, aes(x=price, y=..density.., 
                          color=cut, fill=cut)) + 
  geom_histogram(alpha=.25, bins=30)+ 
  facet_wrap(~cut)
```





---
## More low-carat Ideal cuts?

```{r, fig.width=7, fig.height=5, message=F}
ggplot(data=diamonds, aes(x=carat, y=..density..,
                          color=cut, fill=cut)) +
  geom_histogram(alpha=.25, bins=30) +
  facet_wrap(~cut)
```




---
## All on one using a density plot

```{r, fig.width=7, fig.height=5, message=F}
ggplot(data=diamonds, aes(x=carat, color=cut, fill=cut)) +
  geom_density(alpha=.5)
```





---
## buncha boxplots
```{r}
# discretizing carats into quartiles
carat_levs <- c(0, quantile(diamonds$carat))
diamonds$carat_fac <- factor(cut(diamonds$carat, carat_levs), ordered=TRUE)

ggplot(data=diamonds, aes(x=carat_fac, y=price, color=cut)) + 
  geom_boxplot() 
```



---
## buncha boxplots... better labels (code)

- make better labels using `labs()`

- can save the base plot into an object, and make modifications by adding plot components / specifications


```{r, eval=F}
bp <- ggplot(data=diamonds, aes(x=carat_fac, y=price, color=cut)) + 
  geom_boxplot() 

bp +
  labs(title="price by carat quartile", 
       subtitle="broken down by cut quality",
       x="carat quartiles")
```




---
## buncha boxplots... better labels (plot)
```{r, echo=F}
bp <- ggplot(data=diamonds, aes(x=carat_fac, y=price, color=cut)) + 
  geom_boxplot() 

bp +
  labs(title="price by carat quartile", 
       subtitle="broken down by cut quality",
       x="carat quartiles")
```





---
## Different data:

- we can also specify that different layers use different datasets.

--

- this might be useful if, for example, you generate predictions based on a subset of your data, and want to visualize it against the full data.

```{r}
# sample subset + build loess model
dia_sub <- diamonds[sample(1:nrow(diamonds), 500), ]
mod <- loess(price~carat, data=dia_sub)

# generate predictions
grid <- data.frame(carat = seq(min(diamonds$carat), max(diamonds$carat), length = 100))
grid$price <- predict(mod, newdata = grid)

ggplot(data=diamonds, aes(x=carat, y=price)) + 
  geom_point(alpha=.01) + 
  geom_smooth(method='loess', se=FALSE) +
  geom_line(data=grid, aes(x=carat, y=price), color="hotpink")


ggplot(data=diamonds, aes(y=price, x=table, size=carat, alpha=carat, color=color)) +
  geom_point() +
  facet_wrap(~cut)



```




---
## data format


In general, plotting is easier when your data is clean and in the right shape.  

The `diamonds` dataset has already been processed for use with ggplot2, so plotting with it is much easier than with the datasets you get from your collaborators!

--

`ggplot2` requires dataframes (or, um, "tibbles"... yes.  I know.), and plotting using ggplot2 is much easier when your data 

- is in long format

- factors are ordered in the order you want them displayed

- factors don't have too many levels

- names are cleaned

In the next portion, we'll do some more EDA on the NYTimes dataset, and cover ggplot2 and data prep in a bit more depth.







